---
title: "Cleaning and Wrangling"
format: html
editor: visual
---

```{r}
#| message: false

library(dplyr)
library(tidyr)
library(readr)
```

#Data Sources

#Read in our data

```{r}
catch_original <- read_csv("https://knb.ecoinformatics.org/knb/d1/mn/v2/object/df35b.302.1")
```

## Explore

```{r}
## Prints the column names of my data frame
colnames(catch_original)

## First 6 lines of the data frame
head(catch_original)

## Summary of each column of data
summary(catch_original)

## Prints unique values in a column (in this case, the region)
unique(catch_original$Region)

```

## Cleaning and reshaping

#Using the select() funciton

```{r}
catch_data <- catch_original %>%
    select(Region, Year, Chinook, Sockeye, Coho, Pink, Chum)

head(catch_data)

catch_data <- catch_original %>%
    select(-All,-notesRegCode)
head(catch_data)
```
##check column types
```{r}
glimpse(catch_data)

```

##Replace values in a column with mutate
```{r}
#change chinook to numeric
catch_clean <- catch_data %>%
  mutate(Chinook = as.numeric(Chinook))
```
```{r}
#find NA values
i <- which(is.na(catch_clean$Chinook))
i
catch_data[i,]

catch_clean <- catch_data %>%
    mutate(Chinook = if_else(condition = Chinook == "I", 
                             true = "1", 
                             false = Chinook),
           Chinook = as.numeric(Chinook))

##check
catch_clean[i, ]
```

## Changing shape using pivot_longer and pivot_wider
```{r}


catch_long <- catch_clean %>% 
    #pivot longer all columns except Region and Year
    pivot_longer(
        cols = -c(Region, Year),
        names_to = "species",
        values_to = "catch"
    )

head(catch_long)
```

```{r}
catch_wide <- catch_long %>%
  pivot_wider(names_from = species,
              values_from = catch)
head(catch_wide)
```
##rename columns with rename

```{r}
catch_long <- catch_long %>%
    rename(catch_thousands = catch)

head(catch_long)
```

```{r}
catch_long <- catch_long %>%
    mutate(catch = catch_thousands * 1000) %>%
  select(-catch_thousands)

head(catch_long)
```
## summary statistics
```{r}
mean_region <- catch_long %>%
    group_by(Region) %>%
    summarize(mean_catch = mean(catch))

head(mean_region)

#number of rows for each group
n_region <- catch_long %>%
    group_by(Region) %>%
    summarize(n = n())

head(n_region)
```
## filtering rows with filter()

```{r}
sse_catch <- catch_long %>%
  filter(Region == "SSE")

head(sse_catch)

#filter for over 1million

## Catches over a million fish
catch_million <- catch_long %>%
    filter(catch > 1000000)

## Chinook from SSE data
chinook_sse <- catch_long %>%
    filter(Region == "SSE",
           species == "Chinook")

## OR combine tests with & ("and") or | ("or")... also, we can swap == for %in%
chinook_sse <- catch_long %>%
    filter(Region %in% "SSE" & species %in% "Chinook")
head(chinook_sse)
```

## Sorting data with arrange()

```{r}
mean_region <- catch_long %>%
    group_by(Region) %>%
    summarize(mean_catch = mean(catch)) %>%
    arrange(mean_catch)

head(mean_region)

#mean descending

mean_region <- catch_long %>%
    group_by(Region) %>%
    summarize(mean_catch = mean(catch)) %>%
    arrange(desc(mean_catch))

head(mean_region)
```

## Splitting columns using separate() and unite()
```{r}
sites_df <- data.frame(site = c("HAW-101",
                                "HAW-103",
                                "OAH-320",
                                "OAH-219",
                                "MAU-039"))

sites_df %>%
    separate(site, c("island", "site_number"), "-")

## create `cities_df`
cities_df <- data.frame(city = c("Juneau AK",
                                 "Sitka AK",
                                 "Anchorage AK"))

cities_clean <- cities_df %>%
  separate(city, c("city", "state_code"), " ")
cities_clean
```
```{r}
dates_df <- data.frame(
    year = c("1930",
             "1930",
             "1930"),
    month = c("12",
              "12",
              "12"),
    day = c("14",
            "15",
            "16")
)
dates_df %>%
    unite(date, year, month, day, sep = "-")
```

## All together
```{r}
catch_original <- read_csv("https://knb.ecoinformatics.org/knb/d1/mn/v2/object/df35b.302.1")

mean_region <- catch_original %>%
  select(-All, -notesRegCode) %>% 
  mutate(Chinook = if_else(Chinook == "I", "1", Chinook)) %>% 
  mutate(Chinook = as.numeric(Chinook)) %>% 
  pivot_longer(-c(Region, Year), 
               names_to = "species", 
               values_to = "catch") %>%
  mutate(catch = catch * 1000) %>% 
  group_by(Region) %>% 
  summarize(mean_catch = mean(catch)) %>% 
  arrange(desc(mean_catch))

head(mean_region)
```
#Writing out the results

```{r}
write_csv(mean_region, here::here("data/mean_catch_by_region.csv"))
```


